---
title: "Mildner et al 2017 recapitulation"
author: "Dan Skelly"
date: "Last updated `r Sys.Date()`"
output: html_document
---

```{r}
root_dir <- '~/projects/scRNA-seq/other_data/Mildner_etal_2017_monocytes'
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Seurat)
library(assertthat)
set.seed(1)
knitr::opts_knit$set(root.dir=root_dir)
```

# Read in the data

```{r reading, message=FALSE}
meta <- read_tsv('GSE95701_metadata.txt.gz', skip=13)

xlist <- list()
files <- list.files(pattern="^GSM2522.*gz$", full.names=TRUE)
for (filename in files) {
  message(filename)
  xlist[[filename]] <- read.table(filename)
  #f <- gzfile(filename, 'rb')
  #h <- readLines(f, n=1) %>% strsplit('\t') %>% `[[`(1)
  #close(f)
}
# Check that all genes quantified are the same
for (i in 2:length(files)) {
  assert_that(are_equal(rownames(xlist[[1]]), rownames(xlist[[i]])))
}
# Check for no overlap in cell IDs
cell_ids <- lapply(xlist, colnames)
uids <- unique(unlist(cell_ids))
assert_that(length(uids) == sum(sapply(cell_ids, length)))

# Assemble the final count matrix:
x <- do.call(cbind, unname(xlist))
rownames(x) <- gsub("_", "-", rownames(x))  # in Seurat feature names cannot have _
assert_that(all(meta$well %in% colnames(x)))
```

# Make a Seurat object

From the Mildner et al 2017 paper: Cells with less than 200 UMIs were 
discarded from the analysis. Genes with mean expression smaller than 
0.005 UMIs/cell or with above average expression and low coefficient of 
variance (<1.2) were discarded.

Below, my Seurat commands are loosely based on
`~/repos/cancer/20190703_tumor_metastasis/normalize_and_cluster.r`

```{r seurat_obj, message=FALSE}
metadf <- meta %>% as.data.frame() %>% column_to_rownames('well')
nUMI <- apply(x, 2, sum)
#normdat <- sweep(x, MARGIN=2, STATS=nUMI, FUN="/")
mu <- apply(x, 1, mean)
sigma <- apply(x, 1, sd)
cv <- sigma/mu
ercc <- grepl("^ERCC", rownames(x))
keep <- !ercc & mu >= 0.005 & !is.na(cv) & cv >= 1.2
mono <- CreateSeuratObject(counts=x[keep, nUMI >= 200], project="Mildner2017",
                           meta.data=metadf[nUMI >= 200, ])
mono <- subset(mono, Number_of_cells == 1)
#mono <- PercentageFeatureSet(mono, pattern="^mt-", col.name="percent.mt")
mono <- NormalizeData(mono) %>% FindVariableFeatures() %>% 
  ScaleData(vars.to.regress=c("Amp_batch_ID", "Seq_batch_ID")) %>%
  RunPCA(verbose=FALSE, npcs=50)
```

Do PCA and clustering

```{r pca}
ElbowPlot(mono, ndims=20)
num_pc <- 5
mono <- FindNeighbors(mono, reduction='pca', dims=1:num_pc, verbose=FALSE) %>%
  FindClusters(verbose=FALSE, resolution=0.15) %>%
  RunUMAP(reduction='pca', dims=1:num_pc, verbose=FALSE)
UMAPPlot(mono)
```

# Identifying clusters

```{r id_clusters, fig.width=8, fig.height=7}
VlnPlot(mono, c("Ly6c2", "Itgal", "Cd74", "H2-Aa", "Cd36", "Sell"), ncol=3)
VlnPlot(mono, c('Cd74', 'H2-Aa', 'Ciita', 'Ccr2', 'Cd209a'), ncol=3) # cluster 2/II
```
```{r clusters2, fig.width=8, fig.height=3.5}
VlnPlot(mono, c("Cebpb", "Nr4a1", "Ccr2"), ncol=3)
```
```{r clusters3, fig.width=8, fig.height=7, message=FALSE}
VlnPlot(mono, c('Nr4a1', 'Cebpb', 'Cd36', 'Pparg', 'Itgax', 'Itgal'), ncol=3) # cluster 0/IV
markers <- FindAllMarkers(mono, only.pos=TRUE)
filter(markers, cluster==0) %>% head(5)
filter(markers, cluster==1) %>% head(5)
filter(markers, cluster==2) %>% head(5)
filter(markers, cluster==3) %>% head(5)
```

Cluster identifications are:

 * 0 = Nr4a1, Cebpb, Cd36 -> Ly6C- monocytes, Mildner cluster IV
 * 1 = Ly6c2, Sell -> Ly6C+ monocytes, Mildner cluster I
 * 2 = Cd74, H2-Aa, Cd209a -> Ly6C^{int}, Mildner cluster II
 * 3 = nothing apparent. Just distinguished by Mt2 expression.

# Write Seurat object 

Write an object including just the main monocyte populations to file.

```{r writing}
cnames <- setNames(c("Ly6C- mono", "Ly6C+ mono", "Ly6C^int mono", "Mt2"), 0:3)
mono$cluster <- cnames[mono$seurat_clusters]
mono <- subset(mono, cluster != "Mt2")
filename <- filename <- paste0(root_dir, 
  "/Mildner_etal_2017_monocytes_recapitulate_", 
  format(Sys.time(), "%Y%m%d"), ".RData")
save(mono, file=filename)
```
